---
title: "FC work with the recount2 pancreatic samples"
author: "Sehrish Kanwal"
date: "`r format(Sys.time(), '%a %Y-%b-%d')`"
params:
    dataDir: 
    TumourFile: 
    NormalFile: 
    gene: 
    outDir: 
output: 
  html_document: 
    keep_md: yes
    code_folding: hide
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Required R packages

```{r, message=FALSE}
suppressMessages(library(edgeR))
suppressMessages(library(limma))
suppressMessages(library(Glimma))
suppressMessages(library(gplots))
suppressMessages(library(ggplot2))
suppressMessages(library(dplyr))
suppressMessages(library(reshape))
suppressMessages(library(ggrepel))
options(width = 140)
```

#### Laoding functions

```{r load_function}
##### Assign colours to analysed groups
getTargetsColours <- function(targets) {

    ##### Predefined selection of colours for groups
    targets.colours <- c("red","blue","green","darkgoldenrod","darkred","deepskyblue", "coral", "cornflowerblue", "chartreuse4", "bisque4", "chocolate3", "cadetblue3", "darkslategrey", "lightgoldenrod4", "mediumpurple4", "orangered3","indianred1","blueviolet","darkolivegreen4","darkgoldenrod4","firebrick3","deepskyblue4", "coral3", "dodgerblue1", "chartreuse3", "bisque3", "chocolate4", "cadetblue", "darkslategray4", "lightgoldenrod3", "mediumpurple3", "orangered1")

    f.targets <- factor(targets)
    vec.targets <- targets.colours[1:length(levels(f.targets))]
    targets.colour <- rep(0,length(f.targets))
    for(i in 1:length(f.targets))
    targets.colour[i] <- vec.targets[ f.targets[i]==levels(f.targets)]

    return( list(vec.targets, targets.colour) )
}
```

## Preparing Tumour Data

```{r tumour_data_prep}
seqdata_t <- read.delim(paste(params$dataDir, params$TumourFile, sep="/"), stringsAsFactors = FALSE, comment.char = "#")
dim(seqdata_t)

# New object that contains just the counts
countdata_t <- seqdata_t[,8, drop = FALSE]
head(countdata_t)
dim(countdata_t) 

# Add rownames i.e. GeneIDs to data
rownames(countdata_t) <- seqdata_t[,1]
head(countdata_t)

# Rename column heading for sample
colnames(countdata_t) <- c("Tumour")
head(countdata_t)

# Filtering to remove low expressed genes. The cpm function also adds a small offset to avoid taking log of zero.
myCPM <- cpm(countdata_t)
head(myCPM)

# Which values in myCPM are greater than 0.15?
thresh <- myCPM > 0.15

# This produces a logical matrix with TRUEs and FALSEs
head(thresh)

# Summary of how many TRUEs there are in each row.
table(rowSums(thresh))

# Would like to keep genes that have TRUE thresh value in each row of thresh
keep <- thresh == TRUE

# Subset the rows of countdata to keep the more highly expressed genes
counts_keep_t <- countdata_t[keep, ,  drop = FALSE]

# Converting row names to first coloumn
# counts_keep_t <- tibble::rownames_to_column(counts_keep_t, "gene_id")

# Taking a look at the subset data
summary(keep)
head(counts_keep_t)
dim(counts_keep_t)
str(counts_keep_t)

# Convert counts to DGEList object. First, create a DGEList object. This is an object used by edgeR to store count data. It has a number of slots for storing various parameters about the data.
y <- DGEList(counts_keep_t)

# have a look at y
y

# See what slots are stored in y
names(y)

# Library size information is stored in the samples slot
y$samples

# Get log2 counts per million
logcounts_t <- cpm(y,log=TRUE)

# have a look at logcounts
head(logcounts_t)
dim(logcounts_t)
```

## Preparing Normal Data

```{r normal_data_prep}
# Downloaded pancreatic data from https://jhubiostatistics.shinyapps.io/recount/. Read in the data.
countdata <- read.delim(paste(params$dataDir, params$NormalFile, sep="/"), stringsAsFactors = FALSE, comment.char = "#")
dim(countdata)
head(countdata)

# Rearranging columns to have gene_id in a diff dataframe as that is not a numeric coloumn and the subsequent analysis assumes that the coloumns have numeric values. For now adding gene_ids as rownames.
countdata_n <- countdata[, c(1, 2, 3, 4)]
head(countdata_n)
rownames(countdata_n) <- countdata[,5]

# Filtering to remove low expressed genes. The cpm function also adds a small offset to avoid taking log of zero.
myCPM <- cpm(countdata_n)

# Which values in myCPM are greater than 0.15?
thresh <- myCPM > 0.15

# This produces a logical matrix with TRUEs and FALSEs
head(thresh)

# Summary of how many TRUEs there are in each row.
table(rowSums(thresh))

# Would like to keep genes that have that have at least 2 TRUES in each row of thresh
keep <- rowSums(thresh) >= 2

# Subset the rows of countdata to keep the more highly expressed genes
counts_keep_n <- countdata_n[keep, ,  drop = FALSE]


# Taking a look at the subset data
summary(keep)
head(counts_keep_n)
dim(counts_keep_n)
str(counts_keep_n)

# Convert counts to DGEList object. First, create a DGEList object. This is an object used by edgeR to store count data. It has a number of slots for storing various parameters about the data.
y <- DGEList(counts_keep_n)

# have a look at y
y

# See what slots are stored in y
names(y)

# Library size information is stored in the samples slot
y$samples

# Get log2 counts per million
logcounts_n <- cpm(y,log=TRUE)

# have a look at logcounts
head(logcounts_n)
dim(logcounts_n)
```

## Differential expression analysis

```{r dea}
# Merging count data from tumour sample and normal samples. 
counts_merged <- merge(logcounts_n, logcounts_t, by="row.names")
dim(counts_merged)
head(counts_merged)
rownames(counts_merged)= counts_merged$Row.names
counts_merged <- counts_merged[,2:6]

# Starting the DE analysis. Creating a design matrix first.
design <-  cbind(c(1,1,1,1,0), c(0,0,0,0,1))

# Assigning rownames and column names
rownames(design) <- c("C1", "C2", "C3", "C4", "T")
colnames(design) <- c("WT", "MU")

fit <- lmFit(counts_merged, design)
cont.matrix <- makeContrasts(MUvsWT=MU-WT, levels=design)
fit2 <- contrasts.fit(fit, cont.matrix)
fit2 <- eBayes(fit2)
result <- topTable(fit2, adjust="BH", number = Inf)
```

## Plotting results

### ecdf plot

```{r ecdf_plot, warning=FALSE}

gene <- params$gene

#Preparing quartiles
q1 <- quantile(result$logFC)[2]
q2 <- 0
q3 <- quantile(result$logFC)[4]

#Making a table from quartiles
events <- data_frame(quartiles = c(q1,q2,q3),
                     text = c('Q1','','Q3'))

# Plot stat_ecdf for kallisto 
p <- ggplot(result, aes(result$logFC)) + 
  stat_ecdf(geom = "step", size = 0.25) + 
  geom_vline(data = events, aes(xintercept = events$quartiles), color = "blue", linetype="dotted") +
  geom_text(data = events, mapping = aes(label = events$text, y = 0, x = quartiles), family="Times", size = 3, hjust = 0) +
  geom_point(aes(x=result[gene, 1], y=0.963 , colour = "yellow"), show.legend = FALSE, label=gene) +
  geom_text(aes(x=5.67, label=gene, y=0.93), family="Times", size = 3) +
  coord_cartesian(xlim = c(-5, 10)) +
  labs(title="Empirical Cumulative Density Function", y = "cumulative fraction", x="gene log2 fold change", family="Times")

p
```

## Z-score transformation

Covert expression values into Z-scores, also called standard scores, which indicate how many standard deviations an observation is above or below the mean. These scores are a useful way of putting data from different sources onto the same scale. **Note** that this will not correct for possible batch effects.

```{r z_transform}
counts_merged.z <- as.data.frame(scale(counts_merged))
```

## Comparing cpm and Z-transformed data

### Overall {.tabset}

```{r pca_plots, fig.width = 6, fig.height = 9}
# Load plotly package
suppressMessages(library(plotly))

##### Create directory for output files
pDir <- paste(params$outDir, "PCA", sep = "/")

if ( !file.exists(pDir) ){
  dir.create(pDir, recursive=TRUE)
}

# Assign colours according to defined sample annotation
targets <- c(rep("WT", 4), rep("MU", 1))
targets.colour <- getTargetsColours(targets)


####################
# cpm data
####################

# Perform principal components analysis
expData_pca <- prcomp(t(counts_merged), scale=FALSE)

# Get variance importance for all principal components
importance_pca <- summary(expData_pca)$importance[2,]
importance_pca <- paste(round(100*importance_pca, 2), "%", sep="")

# Generate bar-plot
# Prepare data frame
expData_pca.df <- data.frame(paste0("PC ", c(1:length(expData_pca$sdev))), expData_pca$sdev)
colnames(expData_pca.df) <- c("PC", "Variances")
# The default order will be alphabetized unless specified as below
expData_pca.df$PC <- factor(expData_pca.df$PC, levels = expData_pca.df[["PC"]])

p.pca <- plot_ly(expData_pca.df, x = ~PC, y = ~Variances, type = 'bar', width = 600, height = 400) %>%
  layout(title = "cpm", xaxis = list(title = "Variances captured by principal components"), yaxis = list(title = "% variance explained"), margin = list(l=50, r=50, b=100, t=100, pad=4), autosize = FALSE)

# Save the box-plot as html
htmlwidgets::saveWidget(p.pca, file=paste0(pDir, "/pca.html"), selfcontained = TRUE)


# Generate PCA 2D plot
# Prepare data frame
expData_pca.df <- data.frame(targets, expData_pca$x[,1], expData_pca$x[,2], expData_pca$x[,3])
colnames(expData_pca.df) <- c("Target", "PC1", "PC2", "PC3")
rownames(expData_pca.df) <- colnames(counts_merged)

p.pca_2d <- plot_ly(expData_pca.df, x = ~PC1, y = ~PC2, color = ~Target, text=rownames(expData_pca.df), colors = targets.colour[[1]], type='scatter', mode = "markers", marker = list(size=10, symbol="circle"), width = 500, height = 400) %>%
layout(title = "cpm", xaxis = list(title = paste("PC1 (",importance_pca[1],")",sep="")), yaxis = list(title = paste("PC2 (",importance_pca[2],")",sep="")), margin = list(l=50, r=50, b=50, t=50, pad=4), autosize = FALSE, showlegend = TRUE, legend = list(orientation = "v", y = 0.9))

# Save the box-plot as html
htmlwidgets::saveWidget(p.pca_2d, file=paste0(pDir, "/pca_2d.htm"), selfcontained = TRUE)


# Generate PCA 3D plot
p.pca_3d <- plot_ly(expData_pca.df, x = ~PC1, y = ~PC2, z = ~PC3, color = ~Target, text=rownames(expData_pca.df), colors = targets.colour[[1]], type='scatter3d', mode = "markers", marker = list(size=8, symbol="circle"), width = 400, height = 500) %>%
  layout(title = "cpm", scene = list(xaxis = list(title = paste("PC1 (",importance_pca[1],")",sep="")), yaxis = list(title = paste("PC2 (",importance_pca[2],")",sep="")), zaxis = list(title = paste("PC3 (",importance_pca[3],")",sep="")) ), margin = list(l=50, r=50, b=50, t=50, pad=4), autosize = FALSE, showlegend = TRUE, legend = list(orientation = "v", y = 0.9))

# Save the box-plot as html
htmlwidgets::saveWidget(p.pca_3d, file=paste0(pDir, "/pca_3d.htm"), selfcontained = TRUE)


####################
# Z-transformed data
####################

# Perform principal components analysis
expData_pca <- prcomp(t(counts_merged.z), scale=FALSE)

# Get variance importance for all principal components
importance_pca <- summary(expData_pca)$importance[2,]
importance_pca <- paste(round(100*importance_pca, 2), "%", sep="")

# Generate bar-plot
# Prepare data frame
expData_pca.df <- data.frame(paste0("PC ", c(1:length(expData_pca$sdev))), expData_pca$sdev)
colnames(expData_pca.df) <- c("PC", "Variances")
# The default order will be alphabetized unless specified as below
expData_pca.df$PC <- factor(expData_pca.df$PC, levels = expData_pca.df[["PC"]])

p.pca.z <- plot_ly(expData_pca.df, x = ~PC, y = ~Variances, type = 'bar', width = 600, height = 400) %>%
  layout(title = "Z-transformed", xaxis = list(title = "Variances captured by principal components"), yaxis = list(title = "% variance explained"), margin = list(l=50, r=50, b=100, t=100, pad=4), autosize = FALSE)

# Save the box-plot as html
htmlwidgets::saveWidget(p.pca.z, file=paste0(pDir, "/pca_z.html"), selfcontained = TRUE)


# Generate PCA 2D plot
# Prepare data frame
expData_pca.df <- data.frame(targets, expData_pca$x[,1], expData_pca$x[,2], expData_pca$x[,3])
colnames(expData_pca.df) <- c("Target", "PC1", "PC2", "PC3")
rownames(expData_pca.df) <- colnames(counts_merged.z)

p.pca_2d.z <- plot_ly(expData_pca.df, x = ~PC1, y = ~PC2, color = ~Target, text=rownames(expData_pca.df), colors = targets.colour[[1]], type='scatter', mode = "markers", marker = list(size=10, symbol="circle"), width = 500, height = 400) %>%
layout(title = "Z-transformed", xaxis = list(title = paste("PC1 (",importance_pca[1],")",sep="")), yaxis = list(title = paste("PC2 (",importance_pca[2],")",sep="")), margin = list(l=50, r=50, b=50, t=50, pad=4), autosize = FALSE, showlegend = TRUE, legend = list(orientation = "v", y = 0.9))

# Save the box-plot as html
htmlwidgets::saveWidget(p.pca_2d.z, file=paste0(pDir, "/pca_2d_z.htm"), selfcontained = TRUE)


# Generate PCA 3D plot
p.pca_3d.z <- plot_ly(expData_pca.df, x = ~PC1, y = ~PC2, z = ~PC3, color = ~Target, text=rownames(expData_pca.df), colors = targets.colour[[1]], type='scatter3d', mode = "markers", marker = list(size=8, symbol="circle"), width = 400, height = 500) %>%
  layout(title = "Z-transformed", scene = list(xaxis = list(title = paste("PC1 (",importance_pca[1],")",sep="")), yaxis = list(title = paste("PC2 (",importance_pca[2],")",sep="")), zaxis = list(title = paste("PC3 (",importance_pca[3],")",sep="")) ), margin = list(l=50, r=50, b=50, t=50, pad=4), autosize = FALSE, showlegend = TRUE, legend = list(orientation = "v", y = 0.9))

# Save the box-plot as html
htmlwidgets::saveWidget(p.pca_3d.z, file=paste0(pDir, "/pca_3d_z.htm"), selfcontained = TRUE)

# Detach plotly package. Otherwise it clashes with other graphics devices
detach("package:plotly", unload=FALSE)
```

#### Variance plots

```{r pca_plots_print, fig.width = 6, fig.height = 9}
# Load plotly package
suppressMessages(library(plotly))

# The variances captured by principal components in cpm data
p.pca
# The variances captured by principal components in Z-transformed data
p.pca.z

```

#### PCA plots (2D)

```{r pca_2d_plots_print, fig.width = 6, fig.height = 9}
# PCA 2D plot in cpm data
p.pca_2d
# PCA 2D plot in Z-transformed data
p.pca_2d.z
```

#### PCA plots (3D)

```{r pca_3d_plots_print, fig.width = 6, fig.height = 9}
# "PCA 3D plot in cpm data
p.pca_3d
# cat("PCA 3D plot in Z-transformed data
p.pca_3d.z

# Detach plotly package. Otherwise it clashes with other graphics devices
detach("package:plotly", unload=FALSE)
```

#### Pair-wise comparison heatmaps

```{r pairwise_heatmaps, warning=FALSE, fig.width = 6, fig.height = 9}
# Load plotly packages
suppressMessages(library(plotly))
suppressMessages(library(heatmaply))

##### Create directory for output files
pDir <- paste(params$outDir, "Pairwise_heatmaps", sep = "/")

if ( !file.exists(pDir) ){
  dir.create(pDir, recursive=TRUE)
}


####################
# cpm data
####################

# For overall pair-wise comparison keep 1000 genes with the highest expression variance across samples
rsd<-apply(counts_merged,1,sd)
sel<-order(rsd, decreasing=TRUE)[1:1000]
counts_merged.var <-counts_merged[sel,]

# Calculate the distnace matrix
distance <- as.dist(dist(t(counts_merged.var), method="euclidean"))
plot.top.matrix <- as.matrix(distance)
hc <- hclust(distance, method="ward.D")

# Generate heatmap including the top correlated genes
p.h <- heatmaply(data.frame(plot.top.matrix), Rowv=as.dendrogram(hc), Colv=as.dendrogram(hc), scale_fill_gradient_fun = ggplot2::scale_fill_gradient2(low = "darkred", high = "darkblue", midpoint = mean(plot.top.matrix)), scale="none", trace="none", limits = c(min(plot.top.matrix),max(plot.top.matrix)), hide_colorbar = FALSE, fontsize_row = 8, fontsize_col = 8) %>%
layout(title = "cpm", autosize = TRUE, height = 500, width = 600, margin = list(l=50, r=50, b=50, t=50, pad=4), showlegend = FALSE)

# Save the heatmap as html
htmlwidgets::saveWidget(p.h, file=paste0(pDir, "/pairwise_heatmaps.html"), selfcontained = TRUE)


####################
# Z-transformed data
####################

# For data presentation in the form of heatmap keep 1000 genes with the highest expression variance across samples
rsd<-apply(counts_merged.z,1,sd)
sel<-order(rsd, decreasing=TRUE)[1:1000]
counts_merged.z.var <-counts_merged.z[sel,]

# Calculate the distnace matrix
distance <- as.dist(dist(t(counts_merged.z.var), method="euclidean"))
plot.top.matrix <- as.matrix(distance)
hc <- hclust(distance, method="ward.D")

# Generate heatmap including the top correlated genes
p.h.z <- heatmaply(data.frame(plot.top.matrix), Rowv=as.dendrogram(hc), Colv=as.dendrogram(hc), scale_fill_gradient_fun = ggplot2::scale_fill_gradient2(low = "darkred", high = "darkblue", midpoint = mean(plot.top.matrix)), scale="none", trace="none", limits = c(min(plot.top.matrix),max(plot.top.matrix)), hide_colorbar = FALSE, fontsize_row = 8, fontsize_col = 8) %>%
layout(title = "Z-transformed", autosize = TRUE, height = 500, width = 600, margin = list(l=50, r=50, b=50, t=50, pad=4), showlegend = FALSE)

# Save the heatmap as html
htmlwidgets::saveWidget(p.h.z, file=paste0(pDir, "/pairwise_heatmaps_z.html"), selfcontained = TRUE)

# Detach plotly packages. Otherwise it clashes with other graphics devices
detach("package:heatmaply", unload=FALSE)
detach("package:plotly", unload=FALSE)
```

```{r pairwise_heatmaps_print, warning=FALSE, fig.width = 6, fig.height = 9}
# Load plotly packages
suppressMessages(library(plotly))
suppressMessages(library(heatmaply))

# Pair-wise comparison heatmap for cpm data
p.h
# Pair-wise comparison heatmap for Z-transformed data
p.h.z

# Detach plotly packages. Otherwise it clashes with other graphics devices
detach("package:heatmaply", unload=FALSE)
detach("package:plotly", unload=FALSE)
```

### For selected gene {.tabset}

```{r gene_plots, fig.width = 6, fig.height = 9}
# Load plotly package
suppressMessages(library(plotly))

##### Create directory for output files
pDir_box <- paste(params$outDir, "Boxplots", gene, sep = "/")
pDir_bar <- paste(params$outDir, "Barplots", gene, sep = "/")

if ( !file.exists(pDir_box) ){
  dir.create(pDir_box, recursive=TRUE)
}

if ( !file.exists(pDir_bar) ){
  dir.create(pDir_bar, recursive=TRUE)
}

####################
# cpm
####################

# Extract data for selected gene
counts_merged.gene <- counts_merged[gene, ]

# Order samples accordingly to investigated groups
dataByGroups <- NULL
targetByGroups <- NULL
colourByGroups <- NULL
expr <- list()
averegeExpr <- NULL

for (i in 1:length(unique(targets))) {

    # Separate samples accordingly to investigated groups
    expr.tmp <- counts_merged.gene[ targets %in% unique(sort(targets))[i] ]
    averegeExpr <- c(averegeExpr, rep(mean(as.numeric(expr.tmp)), length(expr.tmp)))
    colour <- targets.colour[[2]][ targets %in% unique(sort(targets))[i] ]

    # Order samples within each group based on the expression level
    expr.tmp <- expr.tmp[order(expr.tmp)]
    colour <- colour[order(expr.tmp)]

    expr[[i]] <- as.numeric(expr.tmp)
    names(expr)[[i]] <- unique(sort(targets))[i]
    dataByGroups <- c(dataByGroups, expr.tmp)
    targetByGroups <- c(targetByGroups, targets[ targets %in% unique(sort(targets))[i] ])
    colourByGroups <- c(colourByGroups, colour)
}

dataByGroups <- unlist(dataByGroups)

# Generate box-plot
# Prepare data frame
gene.expr.df <- data.frame(targets, as.numeric(counts_merged.gene))
colnames(gene.expr.df) <- c("Group", "Expression")

p.box <- plot_ly(gene.expr.df, y= ~Expression, color = ~Group, type = 'box', jitter = 0.3, pointpos = 0, boxpoints = 'all', colors = targets.colour[[1]], width = 400, height = 400) %>%
layout(title = "cpm", yaxis = list( title = paste0(gene, "  expression")), margin = list(l=50, r=50, b=50, t=50, pad=4), autosize = F, legend = list(orientation = 'v', y = 0.5), showlegend=TRUE)

# Save the box-plot as html
htmlwidgets::saveWidget(p.box, file=paste0(pDir_box, "/boxplot.html"), selfcontained = TRUE)


# Generate bar-plot
# Prepare data frame
dataByGroups.df <- data.frame(targetByGroups, names(dataByGroups), as.numeric(dataByGroups))
colnames(dataByGroups.df) <- c("Group","Sample", "Expression")

##### The default order will be alphabetized unless specified as below
dataByGroups.df$Sample <- factor(dataByGroups.df$Sample, levels = dataByGroups.df[["Sample"]])

p.bar <- plot_ly(dataByGroups.df, x = ~Sample, y = ~Expression, color = ~Group, type = 'bar', colors = targets.colour[[1]], width = 600, height = 400) %>%
layout(title = "cpm", xaxis = list( tickfont = list(size = 10), title = ""), yaxis = list(title = paste0(gene, "  expression")), margin = list(l=50, r=50, b=150, t=50, pad=4), autosize = F, legend = list(orientation = 'v', y = 0.5), showlegend=TRUE)

##### Save the bar-plot as html
htmlwidgets::saveWidget(p.bar, file=paste0(pDir_bar, "/barplot.html"), selfcontained = TRUE)


####################
# Z-transformed data
####################

# Extract data for selected gene
counts_merged.z.gene <- counts_merged.z[gene, ]

# Order samples accordingly to investigated groups
dataByGroups <- NULL
targetByGroups <- NULL
colourByGroups <- NULL
expr <- list()
averegeExpr <- NULL

for (i in 1:length(unique(targets))) {

    # Separate samples accordingly to investigated groups
    expr.tmp <- counts_merged.z.gene[ targets %in% unique(sort(targets))[i] ]
    averegeExpr <- c(averegeExpr, rep(mean(as.numeric(expr.tmp)), length(expr.tmp)))
    colour <- targets.colour[[2]][ targets %in% unique(sort(targets))[i] ]

    # Order samples within each group based on the expression level
    expr.tmp <- expr.tmp[order(expr.tmp)]
    colour <- colour[order(expr.tmp)]

    expr[[i]] <- as.numeric(expr.tmp)
    names(expr)[[i]] <- unique(sort(targets))[i]
    dataByGroups <- c(dataByGroups, expr.tmp)
    targetByGroups <- c(targetByGroups, targets[ targets %in% unique(sort(targets))[i] ])
    colourByGroups <- c(colourByGroups, colour)
}

dataByGroups <- unlist(dataByGroups)


# Generate box-plot
# Prepare data frame
gene.expr.df <- data.frame(targets, as.numeric(counts_merged.z.gene))
colnames(gene.expr.df) <- c("Group", "Expression")

p.box.z <- plot_ly(gene.expr.df, y= ~Expression, color = ~Group, type = 'box', jitter = 0.3, pointpos = 0, boxpoints = 'all', colors = targets.colour[[1]], width = 400, height = 400) %>%
layout(title = "Z-transformed", yaxis = list( title = paste0(gene, "  expression (z-score)")), margin = list(l=50, r=50, b=50, t=50, pad=4), autosize = F, legend = list(orientation = 'v', y = 0.5), showlegend=TRUE)

# Save the box-plot as html
htmlwidgets::saveWidget(p.box.z, file=paste0(pDir_box, "/boxplot_z.html"), selfcontained = TRUE)


##### Generate bar-plot
##### Prepare data frame
dataByGroups.df <- data.frame(targetByGroups, names(dataByGroups), as.numeric(dataByGroups))
colnames(dataByGroups.df) <- c("Group","Sample", "Expression")

##### The default order will be alphabetized unless specified as below
dataByGroups.df$Sample <- factor(dataByGroups.df$Sample, levels = dataByGroups.df[["Sample"]])

p.bar.z <- plot_ly(dataByGroups.df, x = ~Sample, y = ~Expression, color = ~Group, type = 'bar', colors = targets.colour[[1]], width = 600, height = 400) %>%
layout(title = "Z-transformed", xaxis = list( tickfont = list(size = 10), title = ""), yaxis = list(title = paste0(gene, "  expression (z-score)")), margin = list(l=50, r=50, b=150, t=50, pad=4), autosize = F, legend = list(orientation = 'v', y = 0.5), showlegend=TRUE)

##### Save the bar-plot as html (PLOTLY)
htmlwidgets::saveWidget(p.bar.z, file=paste0(pDir_bar, "/barplot_z.html"), selfcontained = TRUE)

# Detach plotly package. Otherwise it clashes with other graphics devices
detach("package:plotly", unload=FALSE)
```

#### Box-plots

```{r gene_boxplots_print, fig.width = 6, fig.height = 9}
# Load plotly package
suppressMessages(library(plotly))

# Box-plot using cpm data for selected gene
p.box
# Box-plot using Z-transformed data for selected gene
p.box.z
```

#### Bar-plots

```{r gene_barplots_print, fig.width = 6, fig.height = 9}
# Box-plot using cpm data for selected gene
p.bar
# Box-plot using Z-transformed data for selected gene
p.bar.z

# Detach plotly package. Otherwise it clashes with other graphics devices
detach("package:plotly", unload=FALSE)
```

## To do list

1. Fix mds plots

```{r, eval=FALSE}
sampleinfo <- data.frame(c('ERR674422','ERR674425','ERR674419', 'ERR674424', 'Tumour'), c('pancreatic', 'pancreatic', 'pancreatic', 'pancreatic', 'tumour'), c('normal', 'normal', 'normal', 'normal', 'tumour'))
colnames(sampleinfo) <-  c("SampleName", "CellType", "Status")
levels(sampleinfo$CellType)
col.cell <- c("purple","orange")[sampleinfo$CellType]
data.frame(sampleinfo$CellType,col.cell)

# Redo the MDS with cell type colouring
plotMDS(y,col=col.cell)
# Let's add a legend to the plot so we know which colours correspond to which cell type
legend("topleft",fill=c("purple","orange"),legend=levels(sampleinfo$CellType))
# Add a title
title("Cell type")
```

2. Plot limma log2FC values and difference between groups' mean Z-scores + annotate pancreatic cancer-related genes (*KRAS*, *SMAD4*, *TP53*, *CKDN2A*, etc.)

3. Plot Z-score discributions for investigated groups + annotate pancreatic cancer-related genes

#### Parameters

```{r params_info}
for ( i in 1:length(params) ) {

  cat(paste("Parameter: ", names(params)[i], "\nValue: ", paste(unlist(params[i]), collapse = ","), "\n\n", sep=""))
}
```

#### Session info

```{r session_info}
sessionInfo()
```