---
title: "Investigate expression data distribution for selected genes"
author: "UMCCR"
date: "`r format(Sys.time(), '%d %B, %Y')`"
params:
  exprDir: 
  exprFile: 
  annotFile: 
  transform: 'CPM'
  filter: TRUE
  norm: 'TMM'
  log: TRUE
  genes: 'All'
  results_name: 
  ensembl_version: 86
output:
  html_document:
    keep_md: yes
    code_folding: hide
#    toc: true
#    toc_float: true
---

The aim of this report is investigate the expression data distribution patters of all or user-defined genes in combined expression data (**`r params$exprFile`**) derived from different samples. The data combination includes filtering out genes with low counts `r if ( !params$filter ) { c("(this step is skipped)") }`, transformation (`r params$transform` method, `r if ( !params$log ) { c("with no log-transformation") } else if ( params$log ) { c("followed by log2-transformation") }`), normalisation (`r params$norm` method). The pipeline is based on recommendations from *[RNAseq123](https://master.bioconductor.org/packages/release/workflows/vignettes/RNAseq123/inst/doc/limmaWorkflow.html){target="_blank"}* package.

```{r define_functions, comment=NA, message=FALSE, warning=FALSE}
### Define functions
##### Create 'not in' operator
"%!in%" <- function(x,table) match(x,table, nomatch = 0) == 0
##### Prepare object to write into a file
prepare2write <- function (x) {
  
  x2write <- cbind(rownames(x), x)
  colnames(x2write) <- c("Gene",colnames(x))
  return(x2write)
}
##### Prepare gene data matrix to write into a file
geneMatrix2write <- function (x) {
  
  x2write <- cbind(rownames(x), x)
  colnames(x2write) <- c("Gene",colnames(x))
  return(x2write)
}
##### Assign colours to different groups
getTargetsColours <- function(targets) {
  
##### Predefined selection of colours for groups
targets.colours <- c("red","blue","green","darkgoldenrod","darkred","deepskyblue", "coral", "cornflowerblue", "chartreuse4", "bisque4", "chocolate3", "cadetblue3", "darkslategrey", "lightgoldenrod4", "mediumpurple4", "orangered3","indianred1","blueviolet","darkolivegreen4","darkgoldenrod4","firebrick3","deepskyblue4", "coral3", "dodgerblue1", "chartreuse3", "bisque3", "chocolate4", "cadetblue", "darkslategray4", "lightgoldenrod3", "mediumpurple3", "orangered1")
  f.targets <- factor(targets)
  vec.targets <- targets.colours[1:length(levels(f.targets))]
  targets.colour <- rep(0,length(f.targets))
  for(i in 1:length(f.targets))
    targets.colour[i] <- vec.targets[ f.targets[i]==levels(f.targets)]
  
  return( list(vec.targets, targets.colour) )
}
##### Calculate TPM from RPKM (from http://luisvalesilva.com/datasimple/rna-seq_units.html )
tpm_from_rpkm <- function(x) {
  rpkm.sum <- colSums(x)
  return(t(t(x) / (1e-06 * rpkm.sum)))
}
```

```{r load_libraries, warning=FALSE}
### Load libraries
suppressMessages(library(preprocessCore))
suppressMessages(library(edgeR))
suppressMessages(library(package=paste0("EnsDb.Hsapiens.v", params$ensembl_version), character.only = TRUE))
```

```{r load_data, message=FALSE, warning=FALSE}
##### Read in expression data and associated sample annotation files
data <- read.table(paste(params$exprDir, params$exprFile, sep="/"), sep="\t", as.is=TRUE, header=TRUE, row.names = 1)
targetFile <- read.table(paste(params$exprDir, params$annotFile, sep="/"), sep="\t", as.is=TRUE, header=TRUE)[,c(1:4)]

##### Make sure that there are no duplciated samples in the target file
targetFile <- targetFile[!duplicated(targetFile[,"Sample_name"]),]
rownames(targetFile) <- targetFile[,"Sample_name"]

##### Make syntactically valid names
colnames(data) <- make.names(colnames(data))
rownames(targetFile) <- make.names(rownames(targetFile))

##### Make sure that the target file contains info only about samples present in the data matrix
targetFile <- targetFile[ rownames(targetFile) %in% colnames(data),  ]

##### Make sure that the samples order in the data matrix is the same as in the target file. If the expression matrix contains data for additional samples then write the data subset (containing only samples from the target file) into a file
if ( !all( colnames(data) %in% rownames(targetFile) , na.rm = FALSE)  ) {
  
  data <- data[ , rownames(targetFile) ]

  ##### Save data subset into a file
  write.table(geneMatrix2write(data), file=paste0(normalizePath(params$exprDir), "/", paste0(params$exprFile, ".subset.txt")), sep="\t", quote=FALSE, row.names=FALSE, append = FALSE)
}

##### Subset data to include on user-defined genes
if ( params$genes != "All" || params$genes != "all"  ) {
  
  genes <- unlist(strsplit(params$genes, split=',', fixed=TRUE))
  
  ##### Identify user-defined genes that were not present in the expression matrix
  genes.missing <- genes[ genes %!in% rownames(data) ]

  genes <- genes[ genes %in% rownames(data)  ]
  
  data <- data[ genes,  ]
  
} else {
  
  genes <- rownames(data)
}

##### Write list of missing genes into a file
if ( length(genes) > 0 ) {
    write.table(prepare2write(genes), file = paste0(normalizePath(params$exprDir), "/", paste0(params$results_name, "_", params$transform, "_", params$norm, ".genes.txt")), sep="\t", quote=FALSE, row.names=TRUE, append = FALSE )
}

##### Write list of missing genes into a file
if ( length(genes.missing) > 0 ) {
  
  write.table(prepare2write(genes.missing), file = paste0(normalizePath(params$exprDir), "/", paste0(params$results_name, "_", params$transform, "_", params$norm, ".genes.missing.txt")), sep="\t", quote=FALSE, row.names=TRUE, append = FALSE )
}
```

***

## Library size

Bar-plot illustrating library size for each sample.

```{r library_size_plot, message = FALSE, warning=FALSE, fig.width = 12, fig.height = 9}
suppressMessages(library(plotly))
##### Generate bar-plot for library size. The colours indicate sample groups, as provided in *Target* column in the sample annotation file
##### Assigne colours to targets and datasets
targets.colour <- getTargetsColours(targetFile$Target)
##### Prepare data frame
data.df <- data.frame(targetFile$Target, colnames(data), as.numeric(colSums(data)*1e-6))
colnames(data.df) <- c("Group","Sample", "Library_size")
##### The default order will be alphabetized unless specified as below
data.df$Sample <- factor(data.df$Sample, levels = data.df[["Sample"]])
p <- plot_ly(data.df, x = ~Sample, y = ~Library_size, color = ~Group, type = 'bar', colors = targets.colour[[1]], width = 1000, height = 400) %>%
  layout(title = "", xaxis = list( tickfont = list(size = 10), title = ""), yaxis = list(title = "Library size (millions)"), margin = list(l=50, r=50, b=150, t=50, pad=4), autosize = F, legend = list(orientation = 'v', y = 0.5), showlegend=TRUE)
##### Print htmlwidget
p
##### Save the bar-plot as html (PLOTLY)
#htmlwidgets::saveWidget(as_widget(p), paste0(normalizePath(params$exprDir), "/", paste0(params$results_name, "_RNAseq_libSize.html")), selfcontained = TRUE)
##### Detach plotly package. Otherwise it clashes with other graphics devices
detach("package:plotly", unload=FALSE)
```

## Data transformation and filtering

The read count data is converted into **`r params$transform`**s using *[edgeR](https://bioconductor.org/packages/release/bioc/html/edgeR.html){target="_blank"}* functions. `r if ( !params$filter ) { c("The option for filtering out genes with low counts is switched OFF") } else if ( params$filter ) { c("Genes with low counts were filtered out") }`. `r if ( !params$log ) { c("The data is not log-transformed") } else if ( params$log ) { c("The data is log2-transformed") }`. Plot(s) below present `r params$transform` data distribution.

```{r data_transformation_filtering, comment = NA, message=FALSE, warning=FALSE}
##### Filtering to remove low expressed genes. For differential expression and related analyses, gene expression is rarely considered at the level of raw counts since libraries sequenced at a greater depth will result in higher counts. Rather, it is common practice to transform raw counts onto a scale that accounts for such library size differences. Genes with very low counts across all libraries provide little evidence for differential expression. In the biological point of view, a gene must be expressed at some minimal level before it is likely to be translated into a protein or to be biologically important. In addition, the pronounced discretenes of these counts interferes with some of the statistical approximations that are used later in the pipeline. These genes should be filtered out prior to further analysis. Users should filter with CPM rather than filtering on the counts directly, as the latter does not account for differences in library sizes between samples. For instance for the CPM-transformed data we keep only genes that have CPM of 1
##### Transformation to CPM or TPM scale (see these blogs for details https://www.rna-seqblog.com/rpkm-fpkm-and-tpm-clearly-explained/ and https://haroldpimentel.wordpress.com/2014/05/08/what-the-fpkm-a-review-rna-seq-expression-units/ ).  CPM = Counts Per Million,  TPM = Transcripts Per Kilobase Million. 
##### CPM transformation and filtering
if ( params$filter && params$transform == "CPM" ) {
  
  ##### Create EdgeR DGEList object
  y <- edgeR::DGEList(counts=data,  group=targetFile$Target)
  
  ##### Remove genes with CPM of at least 1 in less than 10% of samples
  #cat("The CPM of 1 (cut-off for removing low expressed genes) corresponds to", round(min(as.numeric(colSums(data)*1e-6)), digits=0), "reads in sample with the lowest sequencing depth, and", round(max(as.numeric(colSums(data)*1e-6)), digits=0), "reads in sample with the greatest sequencing depth\n")
  
  keep <- rowSums(cpm(y)>1) >= ncol(data)/10
  y$filtered <- y[keep, , keep.lib.sizes=FALSE]
  
  #cat(nrow(y$filtered$counts), "genes remained after filtering out of the", nrow(data), "genes in the input read count matrix\n\n")
  
  ##### Calculate normalization factors and transform the raw-scale to CPM. Add small offset to each observation to avoid taking log of zero
  y <- edgeR::calcNormFactors(y)
  y$transformed <- edgeR::cpm(y, normalized.lib.sizes=TRUE, log=params$log, prior.count=0.25)
  y$filtered <- edgeR::calcNormFactors(y$filtered)
  y$filtered.transformed <- edgeR::cpm(y$filtered, normalized.lib.sizes=TRUE, log=params$log, prior.count=0.25)
##### CPM transformation without filtering
} else if ( !params$filter && params$transform == "CPM" ) {
  
  ##### Create EdgeR DGEList object
  y <- edgeR::DGEList(counts=data,  group=targetFile$Target)
  ##### Calculate normalization factors and  and transform the raw-scale to CPM. Add small offset to each observation to avoid taking log of zero
  y <- edgeR::calcNormFactors(y)
  y$transformed <- edgeR::cpm(y, normalized.lib.sizes=TRUE, log=params$log, prior.count=0.25)
  
##### TPM data transformation. We can convert RPKM to TPM in two different ways: from pre-calculated RPKM, by diving by the sum of RPKM values, or directly from the normalized counts. Here we calculate TPM starting from RPKM values computed using edgeR's rpkm function ( from http://luisvalesilva.com/datasimple/rna-seq_units.html )
##### TPM transformation with filtering
} else if ( params$filter && params$transform == "TPM" ) {
  
  ##### Get genes lengths
  edb <- eval(parse(text = paste0("EnsDb.Hsapiens.v", params$ensembl_version)))
  
  gene.length <- lengthOf(edb, filter = GeneIdFilter(rownames(data)))
  
  ##### Check for which genes the lenght info is not available and remove them from the data
  genes.no_length <- rownames(data)[ rownames(data) %!in% names(gene.length)]
  data <- data[ rownames(data) %!in% genes.no_length, ]
  
  ##### Create EdgeR DGEList object
  y <- edgeR::DGEList(counts=data,  group=targetFile$Target)
  
  ##### Calculate normalization factors and convert data into RPKM
  y <- edgeR::calcNormFactors(y)
  y$transformed <- edgeR::rpkm(y, gene.length = gene.length, normalized.lib.sizes=TRUE, log=FALSE)
  
  ##### Remove genes with TPM of at least 0.2 in less than 10% of samples
  keep <- rowSums(y$transformed>0.2) >= ncol(y$transformed)/10
  y$filtered <- y$counts[keep, ]
  y$filtered.transformed <- y$transformed[keep, ]
  
  ##### ... and then to TPM scale. Add small offset to each observation to avoid taking log of zero
  if ( params$log ) {
    
    y$transformed <- log2(tpm_from_rpkm(y$transformed+0.25))
    y$filtered.transformed <- log2(tpm_from_rpkm(y$filtered.transformed+0.25))
    gene.length <- gene.length[ names(gene.length) %in% rownames(y$filtered.transformed) ]
  
  } else {
    
    y$filtered.transformed <- tpm_from_rpkm(y$filtered.transformed)
  }
##### TPM transformation without filtering
} else if ( !params$filter && params$transform == "TPM" ) {
  
  ##### Get genes lengths
  edb <- eval(parse(text = paste0("EnsDb.Hsapiens.v", params$ensembl_version)))
  
  gene.length <- lengthOf(edb, filter = GeneIdFilter(rownames(data)))
  
  ##### Check for which genes the lenght info is not available and remove them from the data
  genes.no_length <- rownames(data)[ rownames(data) %!in% names(gene.length)]
  data <- data[ rownames(data) %!in% genes.no_length, ]
  
  ##### Create EdgeR DGEList object
  y <- edgeR::DGEList(counts=data,  group=targetFile$Target)
  
  ##### Calculate normalization factors and convert data into RPKM
  y <- edgeR::calcNormFactors(y)
  y$transformed <- edgeR::rpkm(y, gene.length = gene.length, normalized.lib.sizes=TRUE, log=FALSE)
  
  ##### ... and then to TPM scale. Add small offset to each observation to avoid taking log of zero
  if ( params$log ) {
    
    y$transformed <- log2(tpm_from_rpkm(y$transformed+0.25))
  
  } else {
    
    y$transformed <- tpm_from_rpkm(y$transformed)
  }
}
```
```{r data_transformation_plot, comment = NA, message=FALSE, warning=FALSE, fig.width = 12, fig.height = 6}
if ( params$filter ) {
  
  par(mfrow=c(1,2))
  
  ##### Before filtering
  plot(density(y$transformed[,1]), lwd=2, ylim=c(0,0.4), las=2, main="", xlab="", col=targets.colour[[2]][1])
  title(main="Transformed data (unfiltered)", xlab=params$transform)
  abline(v=0, lty=3)
  
  for (i in 2:ncol(y$transformed)){
    den <- density(y$transformed[,i])
    lines(den$x, den$y, lwd=2, col=targets.colour[[2]][i])
  }
  legend("topright", legend=levels(factor(targetFile$Target)), fill=targets.colour[[1]], bty="n", bg = "transparent")
  
  ##### After filtering
  plot(density(y$filtered.transformed[,1]), lwd=2, ylim=c(0,0.4), las=2, main="", xlab="", col=targets.colour[[2]][1])
  title(main="Transformed and filtered data", xlab=params$transform)
  abline(v=0, lty=3)
  
  for (i in 2:ncol(y$filtered.transformed)){
    den <- density(y$filtered.transformed[,i])
    lines(den$x, den$y, lwd=2, col=targets.colour[[2]][i])
  }
  legend("topright", legend=levels(factor(targetFile$Target)), fill=targets.colour[[1]], bty="n", bg = "transparent")
  
  ##### Save the plot as pdf file
  pdf(paste0(normalizePath(params$exprDir), "/", paste0(params$results_name, "_", params$transform, "_", params$norm, "_filtering.pdf")), width=8, height=5)
  par(mfrow=c(1,2))
  
  ##### Before filtering
  plot(density(y$transformed[,1]), lwd=2, ylim=c(0,0.4), las=2, main="", xlab="", col=targets.colour[[2]][1])
  title(main="Transformed data (unfiltered)", xlab=params$transform)
  abline(v=0, lty=3)
  
  for (i in 2:ncol(y$transformed)){
    den <- density(y$transformed[,i])
    lines(den$x, den$y, lwd=2, col=targets.colour[[2]][i])
  }
  legend("topright", legend=levels(factor(targetFile$Target)), fill=targets.colour[[1]], bty="n", bg = "transparent")
  
  ##### After filtering
  plot(density(y$filtered.transformed[,1]), lwd=2, ylim=c(0,0.4), las=2, main="", xlab="", col=targets.colour[[2]][1])
  title(main="Transformed and filtered data", xlab=params$transform)
  abline(v=0, lty=3)
  
  for (i in 2:ncol(y$filtered.transformed)){
    den <- density(y$filtered.transformed[,i])
    lines(den$x, den$y, lwd=2, col=targets.colour[[2]][i])
  }
  legend("topright", legend=levels(factor(targetFile$Target)), fill=targets.colour[[1]], bty="n", bg = "transparent")
  invisible(dev.off())
  
} else if ( !params$filter ) {
  
  ##### Without filtering
  plot(density(y$transformed[,1]), lwd=2, ylim=c(0,0.4), las=2, main="", xlab="", col=targets.colour[[2]][1])
  title(main="Transformed data (unfiltered)", xlab=params$transform)
  abline(v=0, lty=3)
  
  for (i in 2:ncol(y$transformed)){
    den <- density(y$transformed[,i])
    lines(den$x, den$y, lwd=2, col=targets.colour[[2]][i])
  }
  legend("topright", legend=levels(factor(targetFile$Target)), fill=targets.colour[[1]], bty="n", bg = "transparent")
  
  ##### Save the plot as pdf file
  pdf(paste0(normalizePath(params$exprDir), "/", paste0(params$results_name, "_", params$transform, "_", params$norm, "_filtering.pdf")), width=8, height=5)
  ##### Without filtering
  plot(density(y$transformed[,1]), lwd=2, ylim=c(0,0.4), las=2, main="", xlab="", col=targets.colour[[2]][1])
  title(main="Transformed data (unfiltered)", xlab=params$transform)
  abline(v=0, lty=3)
  
  for (i in 2:ncol(y$transformed)){
    den <- density(y$transformed[,i])
    lines(den$x, den$y, lwd=2, col=targets.colour[[2]][i])
  }
  legend("topright", legend=levels(factor(targetFile$Target)), fill=targets.colour[[1]], bty="n", bg = "transparent")
  invisible(dev.off())
}
```

## Data normalisation

During the sample preparation or sequencing process, external factors that are not of biological interest can affect the expression of individual samples. For example, samples processed in the first batch of an experiment can have higher expression overall when compared to samples processed in a second batch. It is assumed that all samples should have a similar range and distribution of expression values. Normalisation for sample-specific effects is required to ensure that the expression distributions of each sample are similar across the entire experiment. Normalisation is performed using **`r params$norm`** method.

Box-plots below present `r params$transform` data for individual samples, coloured by sample groups, before and after `r params$norm` normalisation.

```{r data_normalisation_unnorm_vs_norm, comment = NA, message=FALSE, warning=FALSE, fig.width = 12, fig.height = 16}
##### TMM normalsation. Trimmed mean of M-values (https://www.ncbi.nlm.nih.gov/pubmed/20196867) (TMM) is performed using the calcNormFactors function in edgeR. The normalisation factors calculated here are used as a scaling factor for the library sizes. TMM is the recommended for most RNA-Seq data where the majority (more than half) of the genes are believed not differentially expressed between any pair of the samples. It adjusts for RNA composition effect, calculates scaling factors for the library sizes with calcNormFactors function using trimmed mean of M-values (TMM) between each pair of samples. Note, that the raw read counts are used to calculate the normalisation factors
if ( params$transform == "CPM" ) {
  
  ##### Transformations from the raw-scale to CPM and normalisation using TMM method
  if ( params$filter ) {
    
    y$noNorm <-edgeR::cpm(y$filtered, normalized.lib.sizes=FALSE, log=params$log, prior.count=0.25)
    y$filtered.transformed <- edgeR::calcNormFactors(y$filtered, method = params$norm)
    y$norm <- edgeR::cpm(y$filtered.transformed, normalized.lib.sizes=TRUE, log=params$log, prior.count=0.25)
  
  } else {
    
    y$noNorm <- edgeR::cpm(y, normalized.lib.sizes=FALSE, log=params$log, prior.count=0.25)
    y$transformed <- edgeR::calcNormFactors(y, method = params$norm)
    y$norm <- edgeR::cpm(y$transformed, normalized.lib.sizes=TRUE, log=params$log, prior.count=0.25)
  }
  
##### Quantile normalsation (from https://www.biostars.org/p/296992/ )
} else if ( params$transform == "TPM" ) {
  
  ##### Normalisation using quantile method
  if ( params$filter ) {
    
    y$noNorm <- y$filtered.transformed
    y$filtered.transformed <- data.matrix(y$filtered.transformed) 
    y$norm  <- normalize.quantiles(y$filtered.transformed, copy = TRUE)
    colnames(y$norm) <- colnames(y$filtered.transformed)
    rownames(y$norm) <- rownames(y$filtered.transformed)
  
  } else {
    
    y$noNorm <- y$transformed
    y$transformed <- data.matrix(y$transformed) 
    y$norm  <- normalize.quantiles(y$transformed, copy = TRUE)
    colnames(y$norm) <- colnames(y$transformed)
    rownames(y$norm) <- rownames(y$transformed)
  }
}
  
##### Plot expression distribution of samples for unnormalised and normalised data
par(mfrow=c(2,1), mar=c(max(nchar(colnames(y$noNorm)))/2, 5, 3, 2))
  
##### Unnormalised data
boxplot(y$noNorm, las=2, col=targets.colour[[2]], main="", pch=".", las=3)
title(main="Unnormalised data", ylab=params$transform)
legend("topright", legend=levels(factor(targetFile$Target)), fill=targets.colour[[1]], horiz=TRUE, bg = "transparent", box.col="transparent")
  
##### Normalised data
boxplot(y$norm, las=2, col=targets.colour[[2]], main="", pch=".", las=3)
title(main=paste0("Normalised data (", params$norm, ")"), ylab=params$transform)
legend("topright", legend=levels(factor(targetFile$Target)), fill=targets.colour[[1]], horiz=TRUE, bg = "transparent", box.col="transparent")
  
##### Save the plot as pdf file
pdf(paste0(normalizePath(params$exprDir), "/", paste0(params$results_name, "_", params$transform, "_", params$norm, "_normalisation.pdf")), width=8, height=14)
par(mfrow=c(2,1), mar=c(max(nchar(colnames(y$noNorm)))/2, 5, 3, 2))
  
##### Unnormalised data
boxplot(y$noNorm, las=2, col=targets.colour[[2]], main="", pch=".", las=3)
title(main="Unnormalised data", ylab=params$transform)
legend("topright", legend=levels(factor(targetFile$Target)), fill=targets.colour[[1]], horiz=TRUE, bg = "transparent", box.col="transparent")
  
##### Normalised data
boxplot(y$norm, las=2, col=targets.colour[[2]], main="", pch=".", las=3)
title(main=paste0("Normalised data (", params$norm, ")"), ylab=params$transform)
legend("topright", legend=levels(factor(targetFile$Target)), fill=targets.colour[[1]], horiz=TRUE, bg = "transparent", box.col="transparent")
invisible(dev.off())
  
# #######################################################
# ##### Prepare expression data frame for plotting in plotly by adding samples annotation and converting the matrix into two column data frame
# ##### Un-normalised cpm data
# y$transformed.df <- as.data.frame(cbind(rep(rownames(y$transformed), ncol(y$transformed)), rep(colnames(y$transformed), each=nrow(y$transformed)), rep(as.vector(y$filtered$samples$group), each=nrow(y$transformed)), as.numeric(y$transformed)))
# colnames(y$transformed.df) <- c("Gene", "Sample", "Group", "cpm")
# 
# ##### Normalised cpm data
# y$norm.transformed.df <- as.data.frame(cbind(rep(rownames(y$norm.transformed), ncol(y$norm.transformed)), rep(colnames(y$norm.transformed), each=nrow(y$norm.transformed)), rep(as.vector(y$norm$samples$group), each=nrow(y$norm.transformed)), as.numeric(y$norm.transformed)))
# colnames(y$norm.transformed.df) <- c("Gene", "Sample", "Group", "cpm")
# 
# 
# p <- plot_ly(width = 800, height = 600) 
# 
# for ( i in 1:12 ) {
#  p <- add_trace(p, y = y$transformed[,i], type = 'box', name = colnames(y$transformed)[i], jitter = 0.3, pointpos = 0, boxpoints = 'outliers',
#         marker = list(color = 'rgb(9,56,125)'),
#         line = list(color = 'rgb(9,56,125)'),
#         showlegend=FALSE)
#   }
# 
# ##### Save the box-plot as html
# htmlwidgets::saveWidget(as_widget(p), paste0(normalizePath(params$exprDir), "/", paste0(params$results_name, "_", params$transform, "_", params$norm, "_normalisation.html")), selfcontained = TRUE)
########################################################
```

## Expression data distribution patterns


HERE INCLUDE CHUNKS TO INVESTIGATE EXPRESSION DATA DISTRIBUTION PATTERNS FOR SELECTED GENES


***

## Addendum

<details>
<summary>Parameters</summary>
<font size="2">

```{r params_info, comment = NA}
for ( i in 1:length(params) ) {
  cat(paste("Parameter: ", names(params)[i], "\nValue: ", paste(unlist(params[i]), collapse = ","), "\n\n", sep=""))
}
```

</font>
</details>

<details>
<summary>Session info</summary>
<font size="2">

```{r sessioninfo, comment = NA}
devtools::session_info()
```

</font>
</details>
